---
title: "Model fitting"
editor: visual
---

# Set-up

## 

We want to fit a single model to all species, we want this model to test the Giometto hypothesis, that all species can be described by a single body size shape, that is determined by a single parameter, the mean body size of the species.

The lognormal distribution can be written as:

$$
l \sim \mathcal{LN}(\mu, \sigma)
$$

this is equivalent to:

$$
log(l) \sim \mathcal{N}(\mu, \sigma)
$$

we want a distribution of individuals, i, belonging to species, k, can be written as:

$$
l_{k, i} \sim \mathcal{LN}(\mu_k, \sigma_k)
$$

but within each species, we can have spatial, s, and temporal, t, variation in these parameters

$$
l_{ki} \sim \mathcal{LN}(\mu_{kst}, \sigma_{kst})
$$

we can define $\mu_{kst}$ as:

$$
\mu_{kst} = \mu + \beta^\mu_t \mu + \beta^\mu_s \mu + \beta^\mu_k \mu
$$

and $\sigma_{kst}$ as:

$$
\sigma_{kst} = \sigma + \beta^\sigma_t \sigma + \beta^\sigma_s \sigma + \beta^\sigma_k \sigma
$$

similarly we could instead define $\sigma_{kst}$ as a function of $\mu_{kst}$

$$
\sigma_{kst} = (\beta_1 \mu_{kst} + \beta_2) + \beta^{\sigma}_t \sigma + \beta^{\sigma}_s \sigma + \beta^{\sigma}_k \sigma
$$

If Giometto's hypothesis is correct, we would expect $\sigma_{kst}$ could be fully explained by $\mu_{kst}$, i.e.:

$$
\beta_1 \neq 0 \\
\beta_2 \neq 0 \\
\beta^\sigma_t = 0 \\
\beta^\sigma_s = 0 \\
\beta^\sigma_k = 0
$$

```{r}

library(tidyverse)
library(arrow)
library(lubridate)
library(rstan)

```

```{r}

# Observational-level data (body size, abundance, species, survey ID)
obs_data <- 
  "output/data/dat_obs_cleaned.parquet" |> 
  read_parquet()

# Survey-level information
survey_list <- 
  "data/cleaned/survey_list_m1_aus.parquet" |> 
  read_parquet() 

# Reef Life Survey (RLS) size class bins
cutoff <- c(
  2.5, 5, 7.5, 10, 12.5, 15, 20, 25, 30, 35, 40, 50, 
  seq(from = 62.5, to = 200, by = 12.5), 
  seq(from = 250, to = 500, by = 50))


# Body size bin upper and lower bounds
rls_bins <- 
  tibble(
    # size_index = 0:length(cutoff),
    size_class = c(0, cutoff),
    bin_lwr = (size_class + lag(size_class))/2,
    bin_upr = (size_class + lead(size_class))/2
  ) |>
  drop_na()


# Gridding latitude and longitude, to be used later
obs_data_full <- 
  obs_data |>  
  left_join(survey_list, by = "survey_id") |> 
  mutate(lat_grid = floor(latitude), 
         lon_grid = floor(longitude), 
         survey_year = year(survey_date)) 

nspp <- 100

species_subset <- 
  obs_data_full |> 
  count(species_name) |> 
  arrange(desc(n)) |>
  rownames_to_column() |> 
  filter(rowname %in% 30:(30+nspp-1)) |>  
  pull(species_name)

obs_data_subset <- 
  obs_data_full |> 
    filter(species_name %in% species_subset)

loc_table <- 
  obs_data_subset |> 
  select(lat_grid) |> 
  distinct() |> 
  arrange(lat_grid) |> 
  rownames_to_column(var = "loc_index") |> 
  mutate(loc_index = as.numeric(loc_index))

species_table <- 
  obs_data_subset |> 
  select(species_name) |> 
  distinct() |> 
  arrange(species_name) |> 
  rownames_to_column(var = "species_index") |> 
  mutate(species_index = as.numeric(species_index))

year_table <- 
  obs_data_subset |> 
  select(survey_year) |> 
  distinct() |> 
  arrange(survey_year) |> 
  rownames_to_column(var = "year_index") |> 
  mutate(year_index = as.numeric(year_index))

size_table <- 
    obs_data_subset |> 
  select(size_class) |> 
  distinct() |> 
  arrange(size_class) |> 
  rownames_to_column(var = "size_index") |> 
  mutate(size_index = as.numeric(size_index)) |> 
  left_join(rls_bins, by = join_by(size_class))


```

```{r}

stan_data <- 
  obs_data_subset |> 
  count(species_name, size_class, lat_grid, survey_year) |> 
  left_join(year_table, by = join_by(survey_year)) |>
  left_join(size_table, by = join_by(size_class)) |> 
  left_join(loc_table, by = join_by(lat_grid)) |> 
  left_join(species_table, by = join_by(species_name)) |> 
  mutate(lat_grid = scale(lat_grid) |> as.numeric(),
         survey_year = scale(survey_year) |> as.numeric())
  
# Data for stan model
data_list <- list(
  N = nrow(stan_data), 
  K = max(stan_data$size_index),
  L = max(stan_data$loc_index),
  Y = max(stan_data$year_index), 
  S = max(stan_data$species_index), 
  loc_indx = stan_data$loc_index, 
  year_indx = stan_data$year_index, 
  species_indx = stan_data$species_index, 
  loc = stan_data$lat_grid, 
  year = stan_data$survey_year, 
  bin = stan_data$size_index,
  uprsize = size_table$bin_upr, 
  indivs = stan_data$n
)
mod_number <- 10


# model fitting
if(!file.exists(paste0("output/stan/stan_single_mod", mod_number, "_", nspp, "spp.rds"))){
  stan(file = paste0("data/stan_models/lnorm_single_mod_v", mod_number, ".stan"), 
              data = data_list,
              iter = 600, 
              warmup = 200, 
              chains = 3, 
              refresh = 200, 
              seed = 2) |> 
    write_rds(paste0("output/stan/stan_single_mod", mod_number,  "_", nspp, "spp.rds"))
}


xx <- 
  paste0("output/stan/stan_single_mod", mod_number, "_", nspp,  "spp.rds") |> 
  read_rds()

 xx |> 
  traceplot(inc_warmup = TRUE)|> 
  ggsave(filename = paste0("output/figs/stan/stan_single_mod", mod_number,  "_", nspp, "spp.png"), 
         height = 10, 
         width = 15)

 
 plot(xx, 
      show_density = TRUE, 
      ci_level = 0.5, 
      fill_color = "purple" ) |> 
  ggsave(filename = paste0("output/figs/stan/stan_single_mod", mod_number,  "_", nspp, "spp_posteriors.png"), 
         height = 10, 
         width = 15)

```

```{r}

 
 

# version 9 = overall sdlog instead of one per species 
# version 10 = sd_0 is solely a function of meanlog
 
```
