---
title: "Universal body size"
author: "Freddie Heather"
editor: visual
date: "`r format(Sys.time(), '%d %B, %Y')`"

execute:
  freeze: auto  # re-render only when source changes
---

```{r setup}
#| include = FALSE

library(knitr, quietly = TRUE)

knitr::opts_chunk$set(
  error = FALSE,
  warning = FALSE, 
  message = FALSE
)

source("R/data_cleaning.R")

fig_output_path <- \(x) paste0("output/figs/", x, ".png")
stan_input_path <- \(x) paste0("data/stan_models/", x, ".stan")
stan_output_path <- \(x) paste0("output/stan/", x, ".rds")

```

## Background

**Main question:** Is there a universal shape of species body size distributions?

The shape is determined by the probability of being a certain size. I.e. it is the probability distribution that we are interested in.

This can be re-framed by asking the question: Do all species have the same coefficient of variation in their distribution (i.e. sdlog in the lognormal distribution)?

## Analysis

```{r packages}

library(tidyverse, quietly = TRUE)
library(lubridate, quietly = TRUE)
library(rstan, quietly = TRUE)
library(magick)

```

```{r import}

# Observational-level data (body size, abundance, species, survey ID)
# Only selecting non-targeted species
obs_data <- 
  "data/raw/obs_data_m1_aus.parquet" |> 
  read_parquet() |> 
  filter(species_name %in% read_rds("output/data/selected_spp.rds")) |> 
  filter(!(species_name == "Neoodax balteatus" & size_class == 150))

# Survey-level information
survey_list <- 
  "data/cleaned/survey_list_m1_aus.parquet" |> 
  read_parquet() 


# Reef Life Survey (RLS) size class bins
cutoff <- c(
  2.5, 5, 7.5, 10, 12.5, 15, 20, 25, 30, 35, 40, 50, 
  seq(from = 62.5, to = 200, by = 12.5), 
  seq(from = 250, to = 500, by = 50))


# Body size bin upper and lower bounds
rls_bins <- 
  tibble(
    size_index = 0:length(cutoff),
    size_class = c(0, cutoff),
    bin_lwr = (size_class + lag(size_class))/2,
    bin_upr = (size_class + lead(size_class))/2
  ) |>
  filter(size_index != 0)

```

### Single site

Let's look at a single location, across all time, and see the distribution of body sizes.

```{r singlesite-filter}

# "BI-S1" is the site with most surveys
selected_site <- 
  survey_list |> 
  add_count(site_code) |> 
  arrange(desc(n)) |> 
  head(1) |> 
  pull(site_code)


data_filter_site <- 
  obs_data |> 
  left_join(survey_list, by = join_by(survey_id)) |> 
  filter(site_code == selected_site) |> 
  count(species_name, size_class, name = "n_indivs") |> 
  add_count(species_name, 
            wt = n_indivs,
            name = "n_indivs_total") |> 
  mutate(p = n_indivs/n_indivs_total)

```

```{r singlesite-plot-ssd}


data_filter_site |> 
  ggplot() +
  aes(
    x = size_class, 
    y = p
  ) +
  geom_point() + 
  geom_line() +
  facet_wrap(~species_name)

```

A single site does not give you much information. What about a single ecoregion?

### Single ecoregion

```{r singleecor-filter}

# "Bassian" is the ecoregion with most surveys
selected_ecoregion <- 
  survey_list |> 
  add_count(ecoregion) |> 
  arrange(desc(n)) |> 
  head(1) |> 
  pull(ecoregion)


data_filter_ecor <- 
  obs_data |> 
  left_join(survey_list, by = join_by(survey_id)) |> 
  filter(ecoregion == selected_ecoregion) |> 
  count(species_name, size_class, name = "n_indivs") |> 
  add_count(species_name, 
            wt = n_indivs,
            name = "n_indivs_total") |> 
  mutate(p = n_indivs/n_indivs_total)

```

```{r singleecor-plot-ssd}


data_filter_ecor |> 
  ggplot() +
  aes(
    x = size_class, 
    y = p
  ) +
  geom_point() + 
  geom_line() +
  facet_wrap(~species_name)

```

These look more like reasonable distributions.

### Single ecoregion through time

```{r singleecor-byyear-filter}

data_filter_ecor_byyear <- 
  obs_data |> 
  left_join(survey_list, by = join_by(survey_id)) |> 
  filter(ecoregion == selected_ecoregion) |> 
  mutate(year = year(survey_date) |> as.integer()) |> 
  count(species_name, size_class, year,
        name = "n_indivs") |> 
  add_count(species_name, 
            year,
            wt = n_indivs,
            name = "n_indivs_total") |> 
  mutate(p = n_indivs/n_indivs_total) 

```

```{r singleecor-byyear-plot-ssd}


for(yr in sort(unique(data_filter_ecor_byyear$year))){
  
  file_name <- paste0("output/figs/singleecor_byyear/", yr,".png")
  if(!file.exists(file_name)){
      dat <- 
    data_filter_ecor_byyear |> 
    mutate(p = ifelse(year == yr, p, NA))
  
  plot <-  
     dat|> 
    ggplot() +
    aes(
      x = size_class, 
      y = p
    ) +
    geom_point() + 
    geom_path(data = dat |> drop_na()) +
    facet_wrap(~species_name) 
  
  
    ggsave(filename = file_name, 
           plot = plot)
  } 

}



#"output/figs/singleecor_byyear/" |> 
#  list.files(full.names = TRUE) |> 
 # map(image_read) |> 
  #image_join() |> 
  #image_animate(fps = 1) |> 
  #image_write(path = "output/figs/singleecor_byyear/ssd.gif")



# xx2 <- image_join(xx)
# xx3 <- image_animate(xx2, fps = 1)
# image_write(xx3, path = "output/figs/singleecor_byyear/ssd.gif")
```
